---
title: "Using TidyCensus"
author: "MaryJo Webster"
date: "updated July 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


# Getting set up
```{r}

# Step 1: 
# Register for a Census API key: 
# https://api.census.gov/data/key_signup.html


# Step 2:
# Install usethis package (uncomment and then run the next line of code)
# install.packages("usethis")

# Step 3: 
# Load your Census API key into REnviron
# In the Console area type:   usethis::edit_r_environ()

#it will bring up a new script  where you should put this
# (take off the hashtag at the front and insert your API key in the designated spot)

# CENSUS_API_KEY="yourAPIkeygoeshere"



# Step 4:
#Install this package by uncommenting the next line of code (needed for some mapping)
#webshot::install_phantomjs()


# Step 5:
# Close RStudio and restart


#be sure to comment out the install packages sections of this code chunk before you continue running this script
```




```{r, warning=FALSE, message=FALSE}
 packages <- c("tidyverse", "stringr", "janitor", "rmarkdown", "lubridate", "ggthemes", "knitr", "htmltools", "kableExtra", "leaflet", "sf", "mapview", "tidycensus")
if (length(setdiff(packages, rownames(installed.packages()))) > 0) {
  install.packages(setdiff(packages, rownames(installed.packages())), repos = "http://cran.us.r-project.org")  
}


library(tidycensus)#for getting census data via API
library(tidyverse)
library(janitor)
library(stringr)
library(rmarkdown)
library(knitr)
library(leaflet)#for creating maps
library(sf)#simple features 
library(mapview)#for saving image of map
library(purrr)#this is installed as part of tidyverse
options(scipen=999)  #prevents scientific notation


#this loads your census API key from your REnviron file
 census_api_key(Sys.getenv("CENSUS_API_KEY"))

      
      
  #We're going to use an R Package called TidyCensus to interact with the API
      
#Tidycensus
#https://walkerke.github.io/tidycensus/

#Look at the Tidycensus documentation to see the various options
      #key points -- use get_acs() for American Community Survey;
            #use get_decennial() for 2010 or 2020 decennial census results
        #note the syntax for the various geography levels, such as "state" to get state level
      #or "us" to get national data
      
    

    

#We will need to tell the API:
      #which census product (acs, decennial, etc)
      #which year(s)
      #what geography (sometimes this requires picking a state)
      #What variable(s)
      

```

#How to get a list of variables
```{r}
      
      
#the next code chunk below will generate a list of the tables and variables
      #doing this is optional, but sometimes it helps
      #alternatively you can view the variables online. Select the dataset you want from this page: https://www.census.gov/data/developers/data-sets.html.  Then look for "detailed table variables" link
      
#this particular code is asking for the 2019 1-year American community survey data
#it's storing it in a new data frame and then I've set it to write out to csv

    acs_variable_list <- load_variables(2019, "acs1", cache = TRUE)
    
    write.csv(acs_variable_list, 'acs_variable_list_2019.csv', row.names = FALSE)
```


#A simple API call
Median household income for states
```{r, message=FALSE, error=FALSE, warning=FALSE}
 
#We're going to get median household income from table B19013
# from the ACS 1 year data, for 2019
#geography indicates we're getting state level
#because we haven't specified a state, it will give us all states plus D.C. and Puerto Rico


#notice that we are assigning it to a new dataframe
   hhincome_2019 <- get_acs(geography = "state", 
                     year=2019,
                     survey="acs1",
              variables = c(medincome ="B19013_001"))

#notice that I've named my variable "medincome". That's optional
#You could simply put    c("B19013_001")


#view our results
hhincome_2019
```




#Change the geography
```{r}
#Next let's change the geography to get county level
#again, since we're not specifying a state, it gives us all counties in the US

  hhincome_county <- get_acs(geography = "county", 
                     year=2019,
                     survey="acs1",
                     variables = c(medincome = "B19013_001"))
```

#Limit to one state
```{r}
#Now we'll just get the counties from one state
#Note that it will only return the counties that have valid results
#The 1 year ACS only has data for geographies with 65,000 or more people


   hhincome_mn_county <- get_acs(geography = "county", 
                     year=2019,
                     survey="acs1",
                     state="MN",
                     variables = c(medincome = "B19013_001"))

#View results
hhincome_mn_county
```





#Add more variables
```{r}
    hhincome <- get_acs(geography = "state", 
                     year=2019,
                     survey="acs1",
              variables = c(all = "B19013_001",
                            black = "B19013B_001",
                            indian = "B19013C_001",
                            asian = "B19013D_001",
                            other = "B19013F_001",
                            multi = "B19013G_001",
                            white_nh = "B19013H_001",
                            hispanic = "B19013I_001"))

hhincome
```


#Wide vs Tidy output
```{r}
#Notice in the last query, the results came back with one row for each estimate
#meaning there were multiple rows for each state
#That is considered "tidy" output; and that is the default of TidyCensus
#If you want, you can specify wide output
    hhincome_wide <- get_acs(geography = "state", 
                     year=2019,
                     survey="acs1",
              variables = c(all = "B19013_001",
                            black = "B19013B_001",
                            indian = "B19013C_001",
                            asian = "B19013D_001",
                            other = "B19013F_001",
                            multi = "B19013G_001",
                            white_nh = "B19013H_001",
                            hispanic = "B19013I_001"),
              output="wide")

hhincome_wide
#Notice that now we just have 52 rows and the variables are running across in columns
#The ones labeled with an E on the end are estimates; the ones with M are margin of error values

```




#Get multiple years at the same time
```{r}

#The previous queries have only pulled a single year of data at one time
#and they used the most basic syntax for TidyCensus
#Now we're going to use the map_dfr() function from the purrr package to write a loop
#It will loop through the years that we've told it to pull




#This time we'll grab from table B03002 - Hispanic or Latino orgin by race
#we'll also assign our variables to a list of values (note that we reference it in code below)
  my_variables <-  c(totpop = "B03002_001",
                            nothispanic = "B03002_002",
                            nh_whitealone = "B03002_003",
                            nh_black = "B03002_004",
                            nh_indian = "B03002_005",
                            nh_asian = "B03002_006",
                            nh_pacisl = "B03002_007",
                            nh_other = "B03002_008",
                            nh_multi = "B03002_009",
                            hispanic = "B03002_012") 

#Next we'll create a list of the years that we want to pull
    #The API only goes back to 2012
years <- lst(2016, 2017, 2018, 2019)

#Below is the loop code wrapped around the standard TidyCensus syntax
#Note that we're getting state level
#from ACS 1 year
#I've also added a clean_names() function from janitor package 

states_diversity <-  map_dfr(
  years,
  ~ get_acs(
    geography = "state",
    variables = my_variables,
    year = .x,
    survey = "acs1"
    ),
  .id = "year"
  ) %>% clean_names()


#Let's see what we have for Minnesota
states_diversity %>% filter(name=='Minnesota')

```



#Multiple geographies
```{r}
#There are directions below for getting things like "all counties across X number of states" or all tracts in the U.S., but getting a mixed bag of geographies in one swoop won't work. For example, let's say you want the state of Minnesota and then all the counties in the state (and have it in one file for analysis). Only solution I've found is to bring each one down separately (as we did above) and then append them together using bind_rows(), like this....

#GEOID=27 is Minnesota
minnesota <- bind_rows(hhincome_mn_county, hhincome %>% filter(GEOID=='27'))

```

#Multiple places, same geography level
https://mattherman.info/blog/tidycensus-mult/

```{r}
#this is useful if you want to set a particular array of places. This example, grabs all the counties in the upper Midwest states

#make a lookup table with only the state codes you want
my_states <- c("MN", "IA", "WI", "ND", "SD")

#set the variables you want
my_vars <- c(
  total_pop = "B01003_001",
  median_income = "B19013_001"
  )

#pull the data
#Notice that this time I'm asking for the 5-year ACS from 2019
multi_state <- get_acs(
  geography = "county",
  variables = my_vars,
  state = my_states,
  year = 2019,
  survey = "acs5",
  geometry = FALSE
  )

```



#Get all census tracts with purrr
https://walkerke.github.io/2017/05/tidycensus-every-tract/
```{r}

#census tracts is one of the geographies that you can only pull
#from the Census API one state at a time. This is a workaround to get them all in one swoop.

#The census API has a table called fips_codes
#view the table; one record per county
fips_codes

#this code creates a list of all the state postal abbreviations
us <- unique(fips_codes$state)[1:51]

#this uses purrr function to iterate through and get 
#hhincome for all census tracts in every state
tracts_hhincome <- map_df(us, function(x) {
  get_acs(geography = "tract", variables = "B19013_001", 
          state = x, survey="acs5", year=2019)
})



```


#Pull all variables from one or more tables

```{r}

#pull in the variables list for the year and survey you want
acs_variable_list2019 <- load_variables(2019, "acs5", cache = TRUE)

#Tell it which tables to pull from by pulling them from the variable list. Need to have the caret (^) in front of each Table ID number
#this example is pulling from several tables

B1001_vars <- acs_variable_list2019 %>%
  filter(str_detect(name, "^B01001_") | str_detect(name, "^B01001B") | str_detect(name, "^B01001C") |
           str_detect(name, "^B01001D") | str_detect(name, "^B01001F") |str_detect(name, "^B01001G") |
           str_detect(name, "^B01001H") | str_detect(name, "^B01001I")) %>%
  pull(name)


#tell it which years to pull from
years <- lst( 2019)

#this code goes and gets the data from the API
sex_by_age <-  map_dfr(
  years,
  ~ get_acs(
    geography = "state",
    state="MN",
    variables = B1001_vars,
    year = .x,
    survey = "acs1"
  ),
  .id = "year"
) %>% clean_names() 


#Notice that the table doesn't have variables labels (just the numbers)
#Join it back to the variables list to capture the labels
sex_by_age <-  left_join(sex_by_age, acs_variable_list2019 %>% select(name, label), by=c("variable"="name")) %>% 
  mutate(label=case_when(label=='Estimate!!Total'~ 'Total', TRUE ~str_sub(label,18,255)))

                                                           

#Now look at the labels in the new table
#We probably want separate columns for the racial group and  age group

#So here we'll create a new column called racial group and use str_sub from the stringr package
#to look in that label field for the table ID numbers.

sex_by_age <-  sex_by_age %>%
  mutate(racialgroup = case_when(str_sub(variable,1,7)=='B01001B'~'Black',
                                 str_sub(variable,1,7)=='B01001C'~'Am Indian',
                                 str_sub(variable,1,7)=='B01001D'~'Asian',
                                 str_sub(variable,1,7)=='B01001E'~ 'Pac Isl',
                                 str_sub(variable,1,7)=='B01001F'~'Other',
                                 str_sub(variable,1,7)=='B01001G' ~'Multi',
                                 str_sub(variable,1,7)=='B01001H'~ 'White-NonHispanic',
                                 str_sub(variable,1,7)=='B01001I'~'Hispanic',
                                 str_sub(variable,1,7)=='B01001_'~'Total'))

#You can do similar code to create a column with age groups
#Remember we got this data from numerous tables? Some of them used different age groups
#For example, some had a separate field for age 20 and others had a 20 to 24 age group
#the code below accounts for all the variations and puts them in new buckets the way I want it

sex_by_age <-  sex_by_age %>%
  mutate(agegroup = case_when(grepl("Under 5", label) ~ 'Under 5',
         grepl("5 to 9", label)~'5 to 9',
         grepl("10 to 14", label)~'10 to 17',
         grepl("15 to 17", label)~'10 to 17',
         grepl("18 and", label)~'18 to 24',
          grepl("20", label)~'18 to 24',
         grepl("21", label)~'18 to 24',
        grepl("20 to 24", label)~'18 to 24',
        grepl("22 to 24", label)~'18 to 24',
         grepl("25 to", label)~'25 to 34',
         grepl("30 to", label)~'25 to 34',
         grepl("35 to", label)~'35 to 44',
         grepl("40 to", label)~'35 to 44',
         grepl("45 to", label)~'45 to 54',
         grepl("50 to",label)~'45 to 54',
         grepl("55 to", label)~'55 to 64',
           grepl("60 and", label)~'55 to 64',
          grepl("62 to", label)~'55 to 64',
          grepl("65 to", label)~'65 to 74',
          grepl("65 and", label)~'65 to 74',
         grepl("70 to", label)~'65 to 74',
         grepl("67 to", label)~'65 to 74',
         grepl("75 to", label)~'75 to 84',
         grepl("80 to", label)~'75 to 84',
           grepl("85 years",label)~'85 and up',
          TRUE~'Total') )
#The TRUE portion of the syntax assumes that all remaining rows must be the "total" rows that are mixed in

#You may want to do something similar to split out the gender and you'd use a similar approach
#For my purposes, I want to see how the racial diversity differs by age group
#So I'm going to make another field to group the minority race/ethnic groups together with one label

sex_by_age <- sex_by_age %>% mutate(largegroup = case_when(racialgroup=='White-NonHispanic'~'White',
                                                                        racialgroup=='Total'~'Total',
                                                                        TRUE~'People of color'))

#make a new data frame that summarizes groups
sex_by_age_2 <-  sex_by_age %>% filter(largegroup!='Total') %>%  group_by(largegroup, agegroup) %>% summarise(est = sum(estimate))

#Now pivot wider and calculate the percentage
#I also tacked on clean_names so we don't end up with a column named "People of color" (with spaces)
#Notice that agegroup is a character column and it's not arranging it in the way we want
pivot_wider(sex_by_age_2, names_from = largegroup, values_from=est) %>%
  clean_names() %>% 
  mutate(pct_color = people_of_color/(people_of_color+white)) %>%
  arrange(agegroup)

#If you want it to go in the correct age group order, you need to make it a data frame
#then overwrite the agegroup field by converting it to a factor and re-leveling it
sex_by_age_wide <- pivot_wider(sex_by_age_2, names_from = largegroup, values_from=est) %>%
  clean_names() %>% 
  mutate(pct_color = people_of_color/(people_of_color+white))%>%
  mutate(agegroup = fct_relevel(agegroup, c("5 to 9", "10 to 17", "18 to 24", "25 to 34", "35 to 44", "45 to 54", "55 to 64", "65 to 74", "75 to 84", "85 and up")))

#view results, arranging it by the agegroup
sex_by_age_wide %>% arrange(agegroup)
```






#Add geometry to make maps
```{r}
#we're going to switch to the 5-year ACS 
#so we can get data for all counties in Minnesota
#pulling the median household income variable
#Note: geometry = TRUE  (this pulls the TIGER file)

mn_income <- get_acs(geography = "county", 
                     year=2019,
                     survey="acs5",
              variables = c(medincome = "B19013_001"), 
              state = "MN",
              geometry = TRUE) %>%
  clean_names()
```

#Map with ggplot
```{r}
#https://walkerke.github.io/tidycensus/articles/spatial-data.html
#https://www.r-spatial.org/r/2018/10/25/ggplot2-sf.html

#You can make a simple map with ggplot
#See the links above for more details
#we're not going to spend any more time on this other than to show that it's possible


mn_income %>% 
    ggplot(aes(fill = estimate)) + 
  geom_sf(color = NA) + 
  scale_fill_viridis_c(option = "magma") 
```

```{r}
#Export the ggplot map you just made
ggsave("my_ggplot_map.png")
```




#Make an interactive map
```{r}
#tutorial: https://juliasilge.com/blog/using-tidycensus/
#this uses  leaflet package



#this establishes the color palette the map will use
#we're telling it to base the colors on the "estimate" column in our data frame
#Other palettes: https://rstudio.github.io/leaflet/colors.html

pal <- colorNumeric(palette = "viridis", domain = mn_income$estimate)


#st_stransform is projecting onto a particular map

#this code makes the map as a new data frame
my_leaflet_map <- mn_income %>%
    sf::st_transform(crs = "+init=epsg:4326") %>%
    leaflet(width = "100%") %>%
    addProviderTiles(provider = "CartoDB.Positron") %>%
    addPolygons(popup = ~ str_extract(name, "^([^,]*)"),
                stroke = FALSE,
                smoothFactor = 0,
                fillOpacity = 0.7,
                color = ~ pal(estimate)) %>%
    addLegend("bottomright", 
              pal = pal, 
              values = ~ estimate,
              title = "Median household income",
               labFormat = labelFormat(prefix = "$"),
              opacity = 1)

#display the map below
my_leaflet_map

```

#save image of interactive map
```{r}
#save an image of your map using mapshot
#https://www.rdocumentation.org/packages/mapview/versions/2.7.0/topics/mapshot

#this needs that webshot::phantom_js package that you installed above

mapshot(my_leaflet_map, file = "my_leaflet_map.png")


```

#EXport data to a shapefile
```{r}
#uses sf package
#export the data file out as a shapefile for us in ArcGIS/QGIS, etc
#Notice that we're grabbing the underlying data frame, not the map that we made above
st_write(mn_income, "mn_income_map.shp")
```

