---
title: "Tidycensus - NICAR22"
output: html_notebook
---


# Getting set up at home
```{r}

# Step 1: 
# Register for a Census API key: 
# https://api.census.gov/data/key_signup.html


# Step 2:
# Install usethis package (uncomment and then run the next line of code)
# install.packages("usethis")

# Step 3: 
# Load your Census API key into REnviron
# In the Console area type:   usethis::edit_r_environ()

#it will bring up a new script  where you should put this
# (take off the hashtag at the front and insert your API key in the designated spot)

# CENSUS_API_KEY="yourAPIkeygoeshere"




# Step 4:
# Close RStudio and restart


#be sure to comment out the install packages sections of this code chunk before you continue running this script
```




```{r, warning=FALSE, message=FALSE}


library(tidycensus)#for getting census data via API
library(tidyverse)
library(janitor)
library(sf)#for exporting shapefile
library(purrr)#this is installed as part of tidyverse
options(scipen=999)  #prevents scientific notation


############################################
#this loads your census API key from your REnviron file

census_api_key(Sys.getenv("CENSUS_API_KEY"))



      
  #We're going to use an R Package called TidyCensus to interact with the API
      
#Tidycensus
#https://walkerke.github.io/tidycensus/

#Look at the Tidycensus documentation to see the various options
      #key points -- use get_acs() for American Community Survey;
            #use get_decennial() for 2010 or 2020 decennial census results
        #note the syntax for the various geography levels, such as "state" to get state level
      #or "us" to get national data
      
    

    

#We will need to tell the API:
      #which census product (acs, decennial, etc)
      #which year(s)
      #what geography (sometimes this requires picking a state)
      #What variable(s)
      

```

#How to get a list of variables
```{r}
      
      
#the next code chunk below will generate a list of the tables and variables
      #doing this is optional, but sometimes it helps
      #alternatively you can view the variables online. Select the dataset you want from this page: https://www.census.gov/data/developers/data-sets.html.  Then look for "detailed table variables" link
      
#this particular code is asking for the 2019 1-year American community survey data
#it's storing it in a new data frame and then I've set it to write out to csv

    acs_variable_list <- load_variables(2019, "acs1", cache = TRUE)
    
#uncomment the code below to export a list of variables to look at it in Excel
    #write.csv(acs_variable_list, 'acs_variable_list_2019.csv', row.names = FALSE)
```


#A simple API call
Median household income for states
```{r, message=FALSE, error=FALSE, warning=FALSE}
 
#We're going to get median household income from table B19013
# from the ACS 1 year data, for 2019
#geography indicates we're getting state level
#because we haven't specified a state, it will give us all states plus D.C. and Puerto Rico


#notice that we are assigning it to a new dataframe
   hhincome_2019 <- get_acs(geography = "state", 
                     year=2019,
                     survey="acs1",
              variables = c(medincome ="B19013_001"))

#notice that I've named my variable "medincome". That's optional
#You could simply put    c("B19013_001")


#view our results
hhincome_2019
```




#Change the geography
```{r}
#Next let's change the geography to get county level
#again, since we're not specifying a state, it gives us all counties in the US

  hhincome_county <- get_acs(geography = "county", 
                     year=2019,
                     survey="acs1",
                     variables = c(medincome = "B19013_001"))
```

#Limit to one state
```{r}
#Now we'll just get the counties from one state
#Note that it will only return the counties that have valid results
#The 1 year ACS only has data for geographies with 65,000 or more people


   hhincome_mn_county <- get_acs(geography = "county", 
                     year=2019,
                     survey="acs1",
                     state="MN",
                     variables = c(medincome = "B19013_001"))

#View results
hhincome_mn_county
```





#Add more variables
```{r}
    hhincome <- get_acs(geography = "state", 
                     year=2019,
                     survey="acs1",
              variables = c(all = "B19013_001",
                            black = "B19013B_001",
                            indian = "B19013C_001",
                            asian = "B19013D_001",
                            other = "B19013F_001",
                            multi = "B19013G_001",
                            white_nh = "B19013H_001",
                            hispanic = "B19013I_001"))

hhincome
```


#Wide vs Tidy output
```{r}
#Notice in the last query, the results came back with one row for each estimate
#meaning there were multiple rows for each state
#That is considered "tidy" output; and that is the default of TidyCensus
#If you want, you can specify wide output
    hhincome_wide <- get_acs(geography = "state", 
                     year=2019,
                     survey="acs1",
              variables = c(all = "B19013_001",
                            black = "B19013B_001",
                            indian = "B19013C_001",
                            asian = "B19013D_001",
                            other = "B19013F_001",
                            multi = "B19013G_001",
                            white_nh = "B19013H_001",
                            hispanic = "B19013I_001"),
              output="wide")

hhincome_wide
#Notice that now we just have 52 rows and the variables are running across in columns
#The ones labeled with an E on the end are estimates; the ones with M are margin of error values

```




#Get multiple years at the same time
```{r}

#The previous queries have only pulled a single year of data at one time
#and they used the most basic syntax for TidyCensus
#Now we're going to use the map_dfr() function from the purrr package to write a loop
#It will loop through the years that we've told it to pull




#This time we'll grab from table B03002 - Hispanic or Latino orgin by race
#we'll also assign our variables to a list of values (note that we reference it in code below)
  my_variables <-  c(totpop = "B03002_001",
                            nothispanic = "B03002_002",
                            nh_whitealone = "B03002_003",
                            nh_black = "B03002_004",
                            nh_indian = "B03002_005",
                            nh_asian = "B03002_006",
                            nh_pacisl = "B03002_007",
                            nh_other = "B03002_008",
                            nh_multi = "B03002_009",
                            hispanic = "B03002_012") 

#Next we'll create a list of the years that we want to pull
    #The API only goes back to 2012
years <- lst( 2018, 2019)

#Below is the loop code wrapped around the standard TidyCensus syntax
#Note that we're getting state level
#from ACS 1 year
#I've also added a clean_names() function from janitor package 

states_diversity <-  map_dfr(
  years,
  ~ get_acs(
    geography = "state",
    variables = my_variables,
    year = .x,
    survey = "acs1"
    ),
  .id = "year"
  ) %>% clean_names()


#Let's see what we have for Georgia
states_diversity %>% filter(name=='Georgia')

```



# Get decennial data

```{r}

#find the variables that are available 
census2020_list <- load_variables(2020, "pl", cache = TRUE)


#populate a list with the variables you want
race_vars <-  c(totpop ="P2_001N",
                hispanic = "P2_002N",
                whitenh = "P2_005N",
                blacknh = "P2_006N",
                amindnh = "P2_007N",
                asiannh = "P2_008N",
                pacislnh = "P2_009N",
                othernh ="P2_010N",
                tworacenh = "P2_011N")



#make the API call - choose the geography and output
race_eth_state<-  get_decennial(geography = "state", variables = race_vars, year = "2020", output="wide")


```






#Add geometry to make maps
```{r}
#we're going to switch to the 5-year ACS 
#so we can get data for all counties in Minnesota
#pulling the median household income variable
#Note: geometry = TRUE  (this pulls the TIGER file)

mn_income <- get_acs(geography = "county", 
                     year=2019,
                     survey="acs5",
              variables = c(medincome = "B19013_001"), 
              state = "MN",
              geometry = TRUE) %>%
  clean_names()
```

#Map with ggplot
```{r}
#https://walkerke.github.io/tidycensus/articles/spatial-data.html
#https://www.r-spatial.org/r/2018/10/25/ggplot2-sf.html

#You can make a simple map with ggplot
#See the links above for more details
#we're not going to spend any more time on this other than to show that it's possible


mn_income %>% 
    ggplot(aes(fill = estimate)) + 
  geom_sf(color = NA) + 
  scale_fill_viridis_c(option = "magma") 
```

```{r}
#Export the ggplot map you just made
ggsave("my_ggplot_map.png")
```




Make an interactive map
tutorial: https://juliasilge.com/blog/using-tidycensus/
this uses  leaflet package



#EXport data to a shapefile
```{r}
#uses sf package
#export the data file out as a shapefile for us in ArcGIS/QGIS, etc
#Notice that we're grabbing the underlying data frame, not the map that we made above
st_write(mn_income, "mn_income_map.shp")
```

