---
#YAML header
title: "Making HTML pages with RMarkdown"
author: "By MaryJo Webster"
date:  "Last updated: `r Sys.Date()`"
output:
  html_document: 
    toc: true  #This establishes a table of contents
    to_depth: 1  #this indicates there will be just 1 layer
    toc_float: true  #TOC will float as you scroll down
    theme: flatly  #theme for page, http://www.datadreaming.org/post/r-markdown-theme-gallery/
---

#About RMarkdown
This tutorial shows you how to use some features of RMarkdown pages, plus packages such as KableExtra, DT and formattable to beautify your results to share with others in an HTML page.<br>
It is possible to export to PDFs, but that's a bit harder and won't be covered here. A lot of the same concepts apply, though.<br>

Read more:
https://bookdown.org/yihui/rmarkdown/
<br>
RMarkdown tutorial<br>
https://ourcodingclub.github.io/2016/11/24/rmarkdown-1.html

<br>
You'll see there are gray portions of the page and white portions. The gray portions are R code chunks. The white portion is where you can insert text that will be rendered on the page. You don't need to use HTML in this area, but you can add some if you want something specific, like this extra line break I'm going to add here.
<br>
RMarkdown has built-in features for making headers, subheaders and other text. <br>

# A single hashmark makes a main head  <br>
## A double hashmark makes a subhead <br>
Single asterisk puts text in *italics*<br>
A double asterisk puts text in **bold**<br>
<br>
The main header will appear in your table of contents (toc) if the toc depth is set to 1 in the yaml header. The sub-head will show if it's set to 2.<br>


Before we go any further, go to the File menu and choose "New File" and then choose "RMarkdown". It will open another markdown page and you can see what is included on a markdown page when you get started. You'll see that the YAML header is less sophisticated than what I put on this page and you'll see that RStudio has given you some sample code and directions for getting started. You need to keep the YAML header (although you can edit it) and keep the code chunk that includes "knitr::opts_chunk". Everything else can be deleted. That gives you an idea what to expect when you go to make your own markdown page. <br><br>
Back to this page....

```{r setup, include=FALSE}
#this section of code automatically comes when you create an RMarkdown page
#and it's necessary for the page to work properly
#set echo to FALSE so that your code doesn't show up in the HTML page
#warning = FALSE - don't display warnings
#message = FALSE - don't display messages in output
#set figure heights and widths (optional, but a good idea if you are making ggplots)
# cache= False - results are not cached for future renders

knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, cache=FALSE, fig.height = 5, fig.width = 8)
```



```{r import, warning=FALSE, message=FALSE}

#Let's load packages and bring in data to work with in this page


#install packages (including ones for later sessions)
 packages <- c("tidyverse", "stringr", "janitor", "rmarkdown", "lubridate", "ggthemes", "knitr", "htmltools", "kableExtra", "leaflet", "sf", "mapview", "DT", "formattable", "tidycensus")
if (length(setdiff(packages, rownames(installed.packages()))) > 0) {
  install.packages(setdiff(packages, rownames(installed.packages())), repos = "http://cran.us.r-project.org")  
}



#load packages 
library(tidyverse)
library(ggthemes) #this will help make our charts look better
library(janitor) #for data cleanup
library(lubridate) #for working with dates
library(stringr) #for data clean up
library(rmarkdown)
library(knitr)
library(htmltools)
library(kableExtra)
library(DT) #datatable package
library(formattable)
library(tidycensus)


#Import our deaths data
#Note that our data is stored in a sub-directory called "data"

deaths <- read_csv('./data/opiate_deaths.csv',
                   col_types=cols(.default="c",BIRTHDATE=col_date("%m/%d/%Y"),                  DEATHDATE=col_date("%m/%d/%Y"),INJURY_DATE=col_date("%m/%d/%Y"), AGEYEARS="i")) %>% clean_names()




#Import some Census median household income data
#You must have a census API key already loaded for this to work

 census_api_key(Sys.getenv("CENSUS_API_KEY"))

     hhincome <- get_acs(geography = "state", 
                     year=2019,
                     survey="acs1",
              variables = c(all = "B19013_001",
                            black = "B19013B_001",
                            indian = "B19013C_001",
                            asian = "B19013D_001",
                            other = "B19013F_001",
                            multi = "B19013G_001",
                            white_nh = "B19013H_001",
                            hispanic = "B19013I_001"))    

```



# Using KableExtra
The KableExtra package offers some ways to add HTML styling to your output. We'll walk through using it from basic to more advanced.
<br>
https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html

```{r}
#Here we're creating a new dataframe and then using kable to display the whole dataframe
#this is a very basic usage of kable, without adding any styling

under10 <-  deaths %>%  select(firstname, ageyears, rescity, deathdate) %>% 
  filter(ageyears<10) %>% 
  arrange(ageyears)

kable(under10) 



```

## Using Kable on the fly

```{r}

#Here we're just adding it to the end of a query output
deaths %>% 
  summarize( AverageAge = round(mean(ageyears),1) ) %>%
  kable()
```


## Include= false
Use this if you don't want a chunk in your HTML output (it won't even display your results)
```{r, include=FALSE}
deaths %>% 
  summarize( AverageAge = round(mean(ageyears),1) ) %>%
  kable()
```


## Echo =TRUE
Use this if you want the code to display in HTML output, along with your results
```{r, echo=TRUE}

#EVen if you have echo=FALSE set at  the very top of the page, this will over-ride that
#just for this one chunk
deaths %>% 
  summarize( AverageAge = round(mean(ageyears),1) ) %>%
  kable()
```




# Add kable_styling
```{r}
#another line of code will start to put some additional styling
deaths %>% 
  group_by(gender) %>% 
  summarize( Count = n(), AverageAge = round(mean(ageyears),1) ) %>% 
  mutate(Pct = round((Count/sum(Count))*100,1)) %>% 
  select(-Count) %>% 
  kable() %>% 
  kable_styling()
```

# Bootstrap options
Then you can use boostrap options to set the kable styling options such as position, width, responsiveness, etc.

```{r}
deaths %>%   
group_by(gender) %>% 
  summarize( Count = n(), AverageAge = round(mean(ageyears),1) ) %>% #round the age to 1 decimal
  mutate(Pct = round((Count/sum(Count))*100,1)) %>% #create percentage and round it
  select(-Count) %>%   #remove a column from the output
  rename(Gender = gender) %>% #rename a column
  kable() %>% 
kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```

## Other bootstrap options
 kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))<p>
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")<p>
  kable_styling(bootstrap_options = "striped", font_size = 7)



## YOUR TURN
Add some styling

```{r}
deaths %>% 
  group_by(race) %>% 
  summarise(NumberDeaths = n()) %>% 
  arrange(desc(NumberDeaths)) 


```

## YOUR TURN
Write a query and add styling of your choice




# Using DT package
Allows you to make sortable, searchable tables
```{r}
# Number of deaths by county of residence (rescounty), highest to lowest
#Notice that this is just a query with datatable() wrapped around it

datatable(deaths %>% 
  group_by(rescounty) %>% 
  summarize( numdeaths = n() ) %>% 
  arrange( desc(numdeaths) ))


```


Datatable on a data frame<br>
Alternatively, you can make a data frame (or use an existing one) and turn that into a datatable
```{r}

mytable <-  deaths %>% 
  filter(rescounty=='RAMSEY', race!='White') %>% 
  select(firstname, lastname, ageyears, race, deathdate)


datatable(mytable)
```


# Formattable package
This package works similar to kable but also has some styling options for percentages, currency and dates<br>
About: https://renkun-ken.github.io/formattable/ <br>
Tutorial: https://www.displayr.com/formattable/

```{r}
# top_n is from dplyr package; limiting our output to the top 10 (in this case, just to make it cleaner to look at)



formattable(under10) %>% arrange(desc(deathdate)) %>% top_n(10)
```

## Formattable and dates

```{r}
#To format a date, you need to make a new column with mutate
#Notice that then I've added a select line that returns only the columns I want to display

formattable(under10 %>%
              mutate(deathdate2= formattable(under10$deathdate, format="%m/%d/%Y"))  %>%
              select(firstname, ageyears, rescity, deathdate2) ) 
```




## Formattable and Kable
This uses the percent() function from formattable
```{r}
deaths %>%   
group_by(gender) %>% 
  summarize( Count = n(), AverageAge = round(mean(ageyears),1) ) %>% #round the age to 1 decimal
  mutate(Pct = percent(Count/sum(Count), digits=1)) %>% #create percentage and set it to 1 decimal point
  select(-Count) %>%   #remove a column from the output
  rename(Gender = gender) %>% #rename a column to make it look better in the output
  kable() %>% 
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```

## Formattable with currency
```{r}
formattable(hhincome %>%
  filter(variable=='all') %>%
  mutate(hhincome = currency(estimate, digits=0)) %>%
  select(NAME, hhincome) %>%
  arrange(desc(hhincome)))
```

## Color_bar in formattable
```{r}
#Read more about this on the formattable web page linked higher up this page

formattable(deaths %>%
              group_by(gender) %>%
              summarise(count=n()) %>%
              mutate(pct=percent((count/sum(count)))), list(`pct`=color_bar("#FA614B")))



```




# Charts
BEST PRACTICES:<br>
set default sizes at top of page<br>

use themes<br>
https://ggplot2.tidyverse.org/reference/ggtheme.html<br>

https://www.rdocumentation.org/packages/ggthemes/versions/3.5.0  <br>

use scale_x_continuous, scale_y_continuous, scale_x_discrete  for axis styling<br>

use labs() for title and other labels<br>

Get color codes from ColorBrewer: http://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3<br>


```{r}



deaths <-  deaths %>% mutate(deathyr=lubridate::year(deathdate))

ggplot(deaths, aes(x=deathyr)) + 
  geom_bar(color="purple", fill="#3e0045")+
  scale_x_continuous(name="Year", breaks=seq(2005, 2017, 1))+
  theme_hc() +
  labs(title = "Opiate deaths in Minnesota", 
       subtitle = "2005-2017",
       caption = "Source: Minnesota Department of Health",
       x="Year",
       y="Number of deaths")




```



#Knit the page
When your page is ready to go, use the Knit button (blue ball of yarn with knitting needle) on the top of the page to "knit" the page<br>
You need to make sure the knitr package is loaded, otherwise this won't work<br>
You also can't have any errors in your code on the page<br>
It will launch a new window showing you what the output looks like and it will create an html file (with the same name as this file) in your working directory. 