---
title: "Basics of ggplot2"
author: "MaryJo Webster"
date:  "Last updated: `r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```



```{r, warning=FALSE, message=FALSE}
#Load libraries
library(tidyverse) #includes ggplot2
library(janitor) #for cleaning up column names on the import
library(readxl) #for importing excel files
library(scales) #for setting labels on x and y scales
library(lubridate) # for date functions
library(ggthemes)  #to add themes/styling to charts




#import county population data for Massachussetts
ma_pop <-  read_csv('./data/co-est2018-alldata.csv') %>% 
  clean_names() %>% 
  filter(sumlev=='050', stname=='Massachusetts') %>%  #filter to only county-level records for MA
  select(1:18) %>%  #keep only the first 18 columns
  mutate(ctyname = str_trim(str_replace(ctyname, "County", ""))) #get rid of the 'County' as part of the name




```

# Look at our data
```{r}
view(ma_pop)
```

# What's the 2018 population of each county?
```{r}
ma_pop %>%
  select(ctyname, popestimate2018) %>%
  arrange(desc(popestimate2018))
```



# Basics of ggplot2

```{r}

#BASIC CODE FOR A column CHART
#the name of the data frame, and the aesthetics (what to use for x axis and what to use for y axis)
#then on a new line the type of plot (geom_bar, geom_line, geom_point, etc.)
#Notice the plus sign at the end of the first line

ggplot(ma_pop , aes(x=ctyname, y=popestimate2018)) +
  geom_bar(stat="identity")

```

```{r}
#CONVERT TO A HORIZONTAL BAR CHART
ggplot(ma_pop , aes(x=ctyname, y=popestimate2018))+
  geom_bar(stat="identity")+
  coord_flip() #this converts it to horizontal
```



```{r}
#ORDER BY THE POPULATION SIZE - using function from forcats package (part of tidyverse)
#AND ADD SOME COLOR TO THE BARS
#color is for the outline; fill is inside the bars

ggplot(ma_pop, aes(x=fct_reorder(ctyname, popestimate2018), y=popestimate2018)) + 
  geom_bar(stat="identity", color="black", fill="blue")+
  coord_flip()
```


# Let's add more styling


```{r}
#Use colorbrewer2.org to get better colors using hex values

ggplot(ma_pop, aes(x=fct_reorder(ctyname, popestimate2018), y=popestimate2018)) + 
  geom_bar(stat="identity", color="#9ebcda", fill="#9ebcda")+
  coord_flip()
```


# improve the axis titles and labels

```{r}

#https://ggplot2.tidyverse.org/reference/scale_continuous.html
#note: the Y axis is along the bottom because this is flipped



ggplot(ma_pop, aes(x=fct_reorder(ctyname, popestimate2018), y=popestimate2018)) + 
  geom_bar(stat="identity", color="#9ebcda", fill="#9ebcda")+
  coord_flip()+
  scale_y_continuous(name="Population", labels = scales::comma)+ #this uses the scales package
  scale_x_discrete(name="County")
```

```{r}
#Add title/subtitle/caption

ggplot(ma_pop, aes(x=fct_reorder(ctyname, popestimate2018), y=popestimate2018)) + 
  geom_bar(stat="identity", color="#9ebcda", fill="#9ebcda")+
  coord_flip()+
  scale_y_continuous(name="Population", labels = scales::comma)+  
  scale_x_discrete(name=NULL)+       #this tells it to not put a label on the x axis 
  labs(title = "Massachusetts county population", 
       subtitle = "2018",
       caption = "Source: US Census")
```

# Add data labels

```{r}
# this is one of the trickiest things to pull off in ggplot!

ggplot(ma_pop, aes(x=fct_reorder(ctyname, popestimate2018), y=popestimate2018)) + 
  geom_bar(stat="identity", color="#9ebcda", fill="#9ebcda")+
  coord_flip()+
  scale_y_continuous(name="Population", labels = scales::comma)+
  scale_x_discrete(name=NULL)+      
  labs(title = "Massachusetts county population", 
       subtitle = "2018",
       caption = "Source: U.S. Census Bureau")+
  geom_text(stat="identity", label=comma(ma_pop$popestimate2018), hjust= 1, size=3) #this adds labels
```

# Use a theme for styling
```{r}
# use a theme from ggthemes package


#High charts theme
ggplot(ma_pop, aes(x=fct_reorder(ctyname, popestimate2018), y=popestimate2018)) + 
  geom_bar(stat="identity", color="#9ebcda", fill="#9ebcda")+
  coord_flip()+
  scale_y_continuous(name="Population", labels = scales::comma)+
  scale_x_discrete(name=NULL)+      
  labs(title = "Massachusetts county population", 
       subtitle = "2018",
       caption = "Source: U.S. Census Bureau")+
  geom_text(stat="identity", label=comma(ma_pop$popestimate2018), hjust= 1, size=3)+
  theme_hc()
```

```{r}
#FiveThirtyEight theme
ggplot(ma_pop, aes(x=fct_reorder(ctyname, popestimate2018), y=popestimate2018)) + 
  geom_bar(stat="identity", color="#9ebcda", fill="#9ebcda")+
  coord_flip()+
  scale_y_continuous(name="Population", labels = scales::comma)+
  scale_x_discrete(name=NULL)+      
  labs(title = "Massachusetts county population", 
       subtitle = "2018",
       caption = "Source: U.S. Census Bureau")+
  geom_text(stat="identity", label=comma(ma_pop$popestimate2018), hjust= 1, size=3)+
  theme_fivethirtyeight()




```


# Save plot to jpg
```{r}
ggsave("ma_county_pop.jpg", width=8, height=5, units="in", dpi="print")
```




# How have populations in each county changed?



## Calculate percentage change
```{r}
#first we need to create a column with percentage change from 2010 to 2018

ma_pop <- ma_pop %>%
  mutate(pctchange2010_2018 = (popestimate2018-popestimate2010)/popestimate2010)



# look at what we have

ma_pop %>%  
  select(ctyname, pctchange2010_2018) %>% 
  arrange(desc(pctchange2010_2018))
```


# YOUR TURN 
## Make a basic bar chart showing percent change in each county

```{r}





```




# Your Turn
--Add: titles, axis labels, improve colors

hint: you'll need scales::percent instead of scales::comma
```{r}



```




# People who died in deer hunting accidents

```{r}
# import data
deer <- read_xlsx('./data/DEER.xlsx') %>% clean_names()
```


```{r}
# very basic chart
# ggplot assumes you want to count the records since it doesn't have a y value

ggplot(deer, aes(x=injury)) +
  geom_bar()
```



```{r}
# add colors and style the axes

ggplot(deer, aes(x=injury)) +
  geom_bar(color="#7fcdbb", fill="#7fcdbb")+
  scale_y_continuous(name="People", labels = scales::comma)+
  scale_x_discrete(name="Injury severity")
```


# Stacked bar chart
```{r}


deer_by_year <-  deer %>% mutate(yr = year(date)) %>% 
  group_by(yr, injury) %>% 
  summarise(total=n())


ggplot(deer_by_year, aes(x=yr, y=total, fill=injury)) +
  geom_col()+
  scale_fill_manual(values = c("#0868ac", "#43a2ca", "#7bccc4"), labels=c("Fatal", "Major", "Minor"))

```


```{r}
# add labels, etc

ggplot(deer_by_year, aes(x=yr, y=total, fill=injury)) +
  geom_col()+
  scale_fill_manual(values = c("#0868ac", "#43a2ca", "#7bccc4"), labels=c("Fatal", "Major", "Minor"))+
    scale_y_continuous(name="Total")+
  scale_x_continuous(name="Year", breaks=seq(1987, 1994, 1) )+      
  labs(title = "People killed in hunting accidents in Wisconsin", 
       subtitle = "1987-1994",
       caption = "Source: Associated Press")
```


# Line chart- Average age of shooter
```{r}

# here we'll create a column with just the year
# and then we will average the shooter age. We need na.rm=TRUE for it to ignore the NULL values


avg_shooter <-  deer %>% mutate(yr = year(date)) %>% 
  group_by(yr) %>% 
  summarise(avg_age = mean(sage, na.rm=TRUE))



```

```{r}
# basic line chart
ggplot(avg_shooter, aes(x=yr, y=avg_age))+
  geom_line(stat="identity")
```


# Zero base the axis

```{r}
ggplot(avg_shooter, aes(x=yr, y=avg_age))+
  geom_line(stat="identity")+
  scale_y_continuous(name="Average age", limits=c(0,40))+  #zero base the Y axis using limits
  scale_x_continuous(name="Year", breaks=seq(1987, 1994, 1)) 
```



# Pretty up our chart
```{r}
ggplot(avg_shooter, aes(x=yr, y=avg_age))+
  geom_line(stat="identity", color="blue", size=1.25)+
  scale_y_continuous(name="Average age", limits=c(0,40))+  #zero base the Y axis using limits
  scale_x_continuous(name="Year", breaks=seq(1987, 1994, 1)) +
    labs(title = "Average age of people involved in hunting accidents in Wisconsin", 
       subtitle = "1987-1994",
       caption = "Source: Associated Press")+
  theme_hc()
```


## How many incidents each year were self-inflicted versus second-person?
hint: column = type

```{r}
#Make a new data frame first
#leave out the records where the type was NULL/blank

deer_by_type <-  deer %>% mutate(yr = year(date)) %>% 
  filter(type!='NA') %>% 
  group_by(yr, type) %>% 
  summarise(total=n())


ggplot(deer_by_type, aes(x=yr, y=total, fill=type)) +
  geom_col()+
  scale_fill_manual(values = c("#0868ac", "#43a2ca"), labels=c("Self-Inflicted", "Second-Person"))


```

# Ordering your values

```{r}
# how do you figure out what order the labels go in? Check your data table to see what order they appear in
#then also compare your chart back to your data
# Sure-fire approach is to refactor them first
# the order you list them in here is how they will appear in your charts or queries

deer_by_type$type <- factor(deer_by_type$type, levels=c("si", "sp"))
```



# Scatterplot

## Years of experience of shooter and victim

```{r}
ggplot(deer, aes(x=sexper, y=vexper))+
  geom_point(stat="identity")
```

## Why do we have a straight diagonal line of dots?

```{r}
ggplot(deer %>% filter(type=='sp'), aes(x=sexper, y=vexper))+
  geom_point(stat="identity")
```


# YOUR TURN
hint: These will be very similar to earlier ones we did. It's ok to copy code!

## Make a simple bar chart (or column) showing total victims by gender (all years combined)

```{r}

```


## Make a line chart showing total people (regardless of injury severity) each year

```{r}

```



# Knit the page
Push the "Knit" button (near the save button) and it will create an HTML page<br>
It won't work if you have any errors in your code anywhere on the page
