---
title: "Importing and joining"
output: html_notebook
---

```{r}
#data: kindergarten vaccination rates from the 2017-18 schoolyear
#includes all public and private schools serving kindergartners
#source: Minnesota Department of Health
# see record layout in the /data/ directory
#we'll pair this with school data from the Minnesota Department of Education


#use readr for delimited text files and fixed width text file
#https://readr.tidyverse.org/

#Excel
#https://readxl.tidyverse.org/

#Googlesheets (need to have a Gmail account)
#https://datascienceplus.com/how-to-use-googlesheets-to-connect-r-to-google-sheets/

#APIs
#http://www.storybench.org/how-to-access-apis-in-r/

#SPSS, SAS, STATA (2 options here)
#https://haven.tidyverse.org/
#https://cran.r-project.org/web/packages/foreign/index.html

#IPUMS (census microdata)
#https://cran.r-project.org/web/packages/ipumsr/index.html

#muckrakr has function for importing/appending all csv's in the same directory
#https://github.com/andrewbtran/muckrakr

#RMySQL package -- link to a mySQL server


library(tidyverse)
library(janitor)
library(stringr)
library(readxl)
library(scales) #needed for the percent_format on the ggplot
options(scipen=999) #this prevents results from displaying as scientific notation
library(formattable) #we're going to use this to display a percentage in a table




```



```{r}
# BRING IN OUR DATA -------------------------------------------------------------


#import from Excel
#syntax:  df <- read_excel("./data/filename.xlsx", sheet="Sheet1", range="A1:Z500") %>% clean_names()


kvaxx <-  read_excel("./data/Kindergarten_1718.xlsx", sheet="KINDERGARTEN_1718_COPY") %>% clean_names()




#Look at the structure of the data we imported. Use str() function
str(kvaxx)


```




```{r}

#import no response data from csv
#syntax:  df <-  read_csv("./data/filename.csv") %>% clean_names()
#syntax with column formats:  df <-  read_csv("./data/filename.csv", col_types=cols(.default="c")) %>% clean_names()

noresponse <-  read_csv("./data/noresponse_kind_1718.csv", col_types=cols(.default="c")) %>%
  clean_names()

#Let's check how it all came in
str(noresponse)

```






```{r}

# APPENDING DATA FRAMES ------------------------------------------------------

#append no responses to the kindergarten file



#first let's see how well the column names line up. 
#Use names() function on both data frames

#run this chunk and the next chunk, so you can see both sets of results

names(noresponse)



```

```{r}
names(kvaxx)

```


```{r}


#Notice that the noresponse file has a "school_id"
#in the kvaxx file, those numbers are there, but stored in 3 separate fields
#We need to put them together into one field so we can match it to other data
#we'll use mutate to add a new column
#  the paste() function is part of Base R; it's a way to string things together
#if you leave out the sep function (separator), it will put in spaces

kvaxx <-  kvaxx %>% 
  mutate(school_id= paste(district_num, 
                          school_type, 
                          school_num, 
                          sep="-"))
```



```{r}
#here's syntax for appending rows to another file
#the file named first is where the rows from the other file will end up
#note that here we are overwriting the existing dataframe
#we'll use a function from dplyr  called bind_rows()

kvaxx <-  bind_rows(kvaxx, 
                    noresponse)

```



```{r}
# REVIEW WHAT WE HAVE -----------------------------------------------------


#Review our vaccination data
#what can we say from it?
#what questions would we like to ask?
```



# YOUR TURN
```{r}
#Let's bring in another file that will help us answer those questions

#use read_csv() function to import school_list. 

#Set columns to character
#and use clean_names()










```





```{r}

# JOINING DATA FRAMES -----------------------------------------------------

#Now we need to JOIN the school_list to the kvaxx data frame. In other words, we'll add new rows
#Note that the school_list file has far more rows than kvaxx. Any idea why?

#Joining two files together requires having at least one common field. 
#What is that here?

#YOUR TURN
#Use names() function to look at column names of both data frames










```



```{r}
#Syntax for joining 
#left_join means we'll keep all the rows from the 1st table
#notice that I'm making a new dataframe

kvaxx2 <- left_join(kvaxx,
                    school_list,
                    by=c("school_id"="school_id"))
```


```{r}
#Acckk! we added all the fields from school_list, but don't want them all. 
#we can limit what comes over, using dplyr
#You must include the joining field -- in this case, school_id

kvaxx2 <- left_join(kvaxx, 
                    school_list %>% select(school_id, county, schooltype),
                    by=c("school_id"="school_id"))
```




```{r}
# limit to MMR ------------------------------------------------------------


#Let's winnow down to just MMR so we can focus our analysis a bit
#we'll use names() function to see which fields we want to put in a new dataframe
names(kvaxx2)




```


```{r}
#note how we're also going to rename a field on the fly
k_mmr <-  kvaxx2 %>%
  select(school_id, county, schooltype, disname, schname, enroll, complete_mmr,
                                complete_pert_mmr, inprogress_mmr, inprogress_pert_mmr, co_mmr, co_pert_mmr,
                                me_mmr, me_pert_mmr, 
         district_type=school_type)  #this last one renames the school_type field on the fly
```



# Data cleanup
```{r}

#what problems do we need to deal with in this new k_mmr table?



#YOUR TURN
#Let's look at the enroll field
#use dplyr to summarize the enroll field
#arrange descending










```




```{r}

#let's create a new field and populate it depending on what's in the old one
#this uses case_when() function from dplyr
#https://dplyr.tidyverse.org/reference/case_when.html


k_mmr <-  k_mmr %>%  
  mutate(enroll_new = case_when(enroll=='Enrollment < 5' ~ 0, 
                                is.na(enroll) ~ 0,
                                TRUE ~ as.double(enroll)))
```









```{r}
#Look to see if there are any records where enroll_new is NULL
k_mmr %>%
  filter(is.na(enroll_new)) %>%
  select(school_id, enroll, enroll_new)
```







```{r}
#Let's look at school type

k_mmr %>% count(schooltype)

#How about we clean that up, and a little extra in there -- identifying charter schools
#MDE identifies charter schools with the district type="07"

k_mmr %>% count(district_type)

#Use case_when to populate a new field called "schooltype2" that 
#says "charter" for ones where the district_type=="07", 
#says "unknown" for those where schooltype is null   -- is.na(schooltype)
#and transfers the contents of the schooltype field for all others


k_mmr <- k_mmr %>% mutate(schooltype2= case_when(district_type=='07'~'charter',
                                                 is.na(schooltype)~'unknown',
                                                 TRUE~ schooltype))



```


# YOUR TURN
#write a query to group by the schooltype2 column to see what is in there











```{r}
#Another nice thing to have would be a single column indicating if a school 
#doesn't have a high enough vaccination rate
#essentially we're going to put our data into buckets or groups, which will make our analysis easier
#health experts say a "community" needs 90-95% vaccination rate to ensure "herd immunity"
#so we'll say that anything below 90% lacks immunity

k_mmr <- k_mmr %>%  
  mutate(mmr_pocket = case_when(complete_pert_mmr>=.895 ~ 'sufficient',
                                complete_pert_mmr>=0 & complete_pert_mmr<.895 ~ 'insufficient',
                                enroll=='Enrollment < 5' ~ 'Enroll<5',
                                is.na(complete_pert_mmr) ~ 'failed to report',
                                TRUE ~ 'unk'))

```


# YOUR TURN
#How many schools have insufficient vaccination rates? (hint: mmr_pocket)





#How many private, charter and public schools do we have?




# Cross tabs

Kids in "pocket" schools by school type
```{r}
#this group_by table shows us the numbers, but doesn't put it in quite the right format
k_mmr %>% group_by(schooltype2, mmr_pocket) %>%
  summarise(kids = sum(enroll_new))
```



# Pivot wider (crosstab)
```{r}
#We can use pivot_wider to turn it into a crosstab

#there are two ways to do this. One is to first make a dataframe of the grouped results

grouped <-  k_mmr %>% group_by(schooltype2, mmr_pocket) %>%
  summarise(kids = sum(enroll_new))



#then to reference that in the pivot_wider query
pivot_wider(grouped, names_from=mmr_pocket, values_from=kids)

#syntax:
#pivot_wider(df_name, names_from=x, values_from = x)



```

```{r}
#We can also nest the query right into that spot where you put the table name

pockets <-  pivot_wider(k_mmr %>%
              group_by(schooltype2, mmr_pocket) %>%
              summarise(kids = sum(enroll_new)), 
            names_from = mmr_pocket, values_from = kids)


pockets
```





#Percent of students attending "pocket" schools
```{r}
#add a new column that calculates percent insufficient for each school type
#then select only key fields to display

#note this is using the new data frame you created in the code chunk above

#this uses the percent function from formattable package

pockets %>% 
  mutate(pctinsufficient = percent(insufficient/(insufficient+sufficient))) %>%
  select(schooltype2, insufficient, sufficient, pctinsufficient)

```



# Make a chart
```{r}
#chart showing percentage of schools sufficient, insufficient, etc.

#first make a data frame with the data the way we want
mmr_chart <-  k_mmr %>%
  group_by(mmr_pocket) %>%
  summarise(schools=n()) %>%
  mutate(pct = schools/sum(schools))


```

# Basics of a chart
```{r}
ggplot(mmr_chart, aes(x=mmr_pocket, y=pct)) + 
  geom_bar(stat="identity")
```



# Make it pretty
```{r}
#now make a plot
  ggplot(mmr_chart, aes(x = mmr_pocket, y = pct)) + 
  geom_bar(stat = "identity", color="black",fill="#3e0045")+
  scale_y_continuous(labels=percent_format()) +
  labs(title = "MMR vaccination in kindergartens", 
       subtitle = "2017-18",
       caption = "Source: Minnesota Department of Health",
       x="",
       y="Percentage of schools")
```

